MSF_Groupwork-NBA_shotlog

### 2016-10-12 MSFGroupwork-NBAHotHand.Rmd
last part - hot hand

```{r}
###Hot Hand Theory

@@

#Remove 0 from streak_per_player in order to show only the streaks achieved
streak_per_player[[1]] <- streak_per_player[[1]][streak_per_player[[1]] != 0]
for(i in 1:length(streak_per_player)) {
  streak_per_player[[i]] <- streak_per_player[[i]][streak_per_player[[i]] != 0]
}
streak_per_player

#Create a data frame for the strikes of each one of the 3 players (cannot create a data frame including all of them, because each list has different length)
player1_df <- data.frame('streaks' = streak_per_player[[1]])
player2_df <- data.frame('streaks' = streak_per_player[[2]])
player3_df <- data.frame('streaks' = streak_per_player[[3]])

#Show distribution for each player's streak lengths
gg1 <- ggplot(data = player1_df, aes(x = streaks)) + geom_histogram(binwidth = 1) + labs(title = ourtable[1,'player_name'])
gg2 <- ggplot(data = player2_df, aes(x = streaks)) + geom_histogram(binwidth = 1) + labs(title = ourtable[2,'player_name'])
gg3 <- ggplot(data = player3_df, aes(x = streaks)) + geom_histogram(binwidth = 1) + labs(title = ourtable[3,'player_name'])
grid.arrange(gg1, gg2, gg3, ncol = 3)

#Create a boxplot for each player's streak lengths
boxplot(player1_df$streaks) 
summary(player1_df$streaks) #Check that boxplot is valid
IQR(player1_df$streaks) #Check that boxplot is valid
#The typical length of a streak is 1 (median = 1) - The Interquartile Range is 1 - Streak length of 6 is unusually high compared to the rest of the distribution
boxplot(player2_df$streaks)
summary(player2_df$streaks)
IQR(player2_df$streaks)
#The typical length of a streak is 1.5 (median = 1.5) - The Interquartile Range is 1 - Streak length of 4 is unusually high compared to the rest of the distribution
boxplot(player3_df$streaks)
summary(player3_df$streaks)
IQR(player3_df$streaks)
#The typical length of a streak is 8 (median = 8) - The Interquartile Range is 5 - There is a streak lengths of 13(!!!)

#We can see that the distribution of all players is right skewed and that all of them (especially player3) have some long shooting streaks. To prove/disprove the hot hand theory we need to examine the indipendence of the shots. If each shot that a player takes is independent of the next shot, then a missed/made shot will not affect the propability that the next shot will be missed/made. So, if consecutive shots are independent, we disprove hot hand theory, while if consecutive shots are dependent, we prove the hot hand theory.

#Find the shooting percentage for each of the 3 players
j <- 1
shooting_percentage <- list()
for(i in best_players_df$player_id) {
  shooting_percentage[j] <- sum(df[df[ ,'player_id'] == i, 'FGM'])/ nrow(df[df[ ,'player_id'] == i, ])
  j <- j+1
}
shooting_percentage

###PLAYER 1 VS SAMPLE PLAYER###
#Simulation - create a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter1 
set.seed(974)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[1]]), replace = TRUE, prob = c(shooting_percentage[[1]], 1 - shooting_percentage[[1]]))
simul_shooter

@@

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar() + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar() + labs(title = 'Kyrie Irving')
grid.arrange(g1, g2, ncol = 2)
#We can see that streak length distributions are quite similar, so we can conclude that Kyrie Irving doesn't have a hot hand.

###PLAYER 2 VS SAMPLE PLAYER###
#Simulation - create a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter1 
set.seed(572)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[2]]), replace = TRUE, prob = c(shooting_percentage[[2]], 1 - shooting_percentage[[2]]))
simul_shooter

@@

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar() + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar() + labs(title = 'Mo Williams')
grid.arrange(g1, g2, ncol = 2)

###PLAYER 3 VS SAMPLE PLAYER###
#Simulation - create a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter1 
set.seed(57)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[3]]), replace = TRUE, prob = c(shooting_percentage[[3]], 1 - shooting_percentage[[3]]))
simul_shooter

@@

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar() + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar() + labs(title = 'Klay Thompson')
grid.arrange(g1, g2, ncol = 2)
#Klay Thompson is an exception, had 13 streaks in a row! We cannot find a similar distribution in our sample, so we can assume that he is likely to have a hot hand. Fearther statistical analysis (logistic regression) should be done in order to reach a conclusion.
```

>>>>

##Hot Hand Theory
###Introduction
The analysis aims to show the effect of the 'hot hand theory' on NBA players. The analysis is based on the hypothesis that a shooter with a hot hand will have shots that are not independent of one another. If the shooter makes his first shot, the probability of making his next shot is higher.

###Methods
The steps to approximate this question are as follows:
1. Calculate streaks achieved in all games and from all players
2. Categorize streaks per player and game, focusing on the 3 players with the top 3 performances in terms of points
3. Extract zeros and keep only streaks above 1
4. Visualise streaks' length distribution - barrplots
5. Visualise streaks' length via boxplot
6. Calculate the shooting percentage of the 3 players 
7. Create a sample independent shooter setting as shooting percentage the real player's shooting percentage
8. Compare distributions of sample and real player

```{r}

@@

#Remove 0 from streak_per_player in order to show only the streaks achieved
streak_per_player[[1]] <- streak_per_player[[1]][streak_per_player[[1]] != 0]
for(i in 1:length(streak_per_player)) {
  streak_per_player[[i]] <- streak_per_player[[i]][streak_per_player[[i]] != 0]
}
streak_per_player
```

###Barplots
```{r}

#Create a data frame for the strikes of each one of the 3 players (cannot create a data frame including all of them, because each list has different length)
player1_df <- data.frame('streaks' = streak_per_player[[1]])
player2_df <- data.frame('streaks' = streak_per_player[[2]])
player3_df <- data.frame('streaks' = streak_per_player[[3]])
best_players_df
#Show distribution for each player's streak lengths
gg1 <- ggplot(data = player1_df, aes(x = streaks)) + geom_bar(fill = 'brown') + labs(title = 'Kyrie Irnving')
gg2 <- ggplot(data = player2_df, aes(x = streaks)) + geom_bar(fill = 'red') + labs(title = 'Mo Williams')
gg3 <- ggplot(data = player3_df, aes(x = streaks)) + geom_bar(fill = 'blue') + labs(title = 'Klay Thompson')
grid.arrange(gg1, gg2, gg3, ncol = 3)
```

The distribution of the three players is unimodal and left skewed. There are shown some extremely long shooting streaks made(i.e. Klay Thompson with 13 shots in a row).

###Boxplots
####Boxplot for Kyrie Irnving
```{r}
#Create a boxplot for each player's streak lengths 
boxplot(player1_df$streaks) 
summary(player1_df$streaks) #Check that boxplot is valid
IQR(player1_df$streaks) #Check that boxplot is valid
```

1. The typical length of a streak is 1 (median = 1)
2. The Interquartile Range is 1 
3. Streak length of 6 is unusually high compared to the rest of the distribution

####Boxplot for Mo Williams
```{r}
boxplot(player2_df$streaks)
summary(player2_df$streaks)
IQR(player2_df$streaks)
```

1. The typical length of a streak is 1.5 (median = 1.5)
2. The Interquartile Range is 1
3. Streak length of 4 is unusually high compared to the rest of the distribution

####Boxplot for Klay Thompson
```{r}
boxplot(player3_df$streaks)
summary(player3_df$streaks)
IQR(player3_df$streaks)
```

1. The typical length of a streak is 8 (median = 8)
2. The Interquartile Range is 5
3. Streak length of 13 is unsually high compared to the rest of the distribution

### Plots Result

The distribution of all players is left skewed and all of them (especially player3) have some long shooting streaks. In order to prove/disprove the hot hand theory,  the examination of the indipendence of the shots is required. If each shot that a player takes is independent of the next shot, then the player has the same probability of hitting each shot regardless of the previous shot. If each shot that a player takes is dependent of the next shot, then the probability of making the next shot is higher - hot hand. Sampling an independent shooter having the same shooting percentage with the real one and comparing his distribution with real player's distribution gives further insight in the specific aera.

```{r}
#Find the shooting percentage for each of the 3 players
j <- 1
shooting_percentage <- list()
for(i in best_players_df$player_id) {
  shooting_percentage[j] <- sum(df[df[ ,'player_id'] == i, 'FGM'])/ nrow(df[df[ ,'player_id'] == i, ])
  j <- j+1
}
shooting_percentage
```

###Kyrie Irving
```{r}
###PLAYER 1 VS SAMPLE PLAYER###
#Simulation - create a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter1 
set.seed(974)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[1]]), replace = TRUE, prob = c(shooting_percentage[[1]], 1 - shooting_percentage[[1]]))
simul_shooter

@@

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar() + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar() + labs(title = 'Kyrie Irving')
grid.arrange(g1, g2, ncol = 2
```

The streak length distributions are quite similar, so we can conclude that Kyrie Irving doesn't have a hot hand.

### Mo Williams
```{r}
###PLAYER 2 VS SAMPLE PLAYER###
#Simulation - create a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter1 
set.seed(572)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[2]]), replace = TRUE, prob = c(shooting_percentage[[2]], 1 - shooting_percentage[[2]]))
simul_shooter

@@

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar() + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar() + labs(title = 'Mo Williams')
grid.arrange(g1, g2, ncol = 2)
```

The streak length distributions are quite similar, so we can conclude that Mo Williams doesn't have a hot hand.

###Klay Thompson
```{r}
###PLAYER 3 VS SAMPLE PLAYER###
#Simulation - create a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter1 
set.seed(57)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[3]]), replace = TRUE, prob = c(shooting_percentage[[3]], 1 - shooting_percentage[[3]]))
simul_shooter

@@

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar() + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar() + labs(title = 'Klay Thompson')
grid.arrange(g1, g2, ncol = 2)
```

Klay Thompson is an exception making 13 shots in a row. There is no such distribution in our sample, so it coulb be assumed that he is likely to have a hot hand. Further statistical analysis (logistic regression) should be done in order to reach a conclusion.
